Discrete Quantum Theories and Computing

### 2018-03-14 dissertation.tex
dissertation.lyx Updating...

\begin{comment}
An $n$-qubit pure state in the conventional quantum theory (CQT)
is represented as a vector in the $2^{n}$-dimensional Hilbert space.
If we eliminate all symmetries, an irreducible state is actually a
point in the projective Hilbert space \cite{MosseriDandoloff2001,Jaeger2007}
also known as the complex projective space~$\CP{2^{n}-1}$ \cite{Hatcher2001,Bengtsson2007}.
These irreducible quantum states can also be classified as product
states and entangled states, and the latter one plays the essential
role for pure-state algorithms \cite{Mermin2007,Jaeger2007}.
\end{comment}
It has been 36 years since Feynman~\cite{Feynman1982Simulating}
proposed the idea about quantum computation. Although there are many
models for quantum computing, the circuit model is still one of the
most widespread models~\cite{544199} no matter whether in theoretical
study or in building a real quantum computer. Information in a quantum
circuit is stored in qubits. Each qubit is a two-dimensional quantum
system. For a two-qubit system, if the first and second qubit are
described as the state~$\ket{\psi}=\alpha_{0}\ket{0}+\alpha_{1}\ket{1}$
and $\ket{\phi}=\beta_{0}\ket{0}+\beta_{1}\ket{1}$, the state of
the whole system is their tensor product 
\begin{equation}
\begin{aligned} & \ket{\psi}\ket{\phi}=\ket{\psi}\otimes\ket{\phi}=\left(\alpha_{0}\ket{0}+\alpha_{1}\ket{1}\right)\otimes\left(\beta_{0}\ket{0}+\beta_{1}\ket{1}\right)\\
={} & \alpha_{0}\beta_{0}\ket{0}\otimes\ket{0}+\alpha_{0}\beta_{1}\ket{0}\otimes\ket{1}+\alpha_{1}\beta_{0}\ket{1}\otimes\ket{0}+\alpha_{1}\beta_{1}\ket{1}\otimes\ket{1}\\
={} & \alpha_{0}\beta_{0}\ket{0}\ket{0}+\alpha_{0}\beta_{1}\ket{0}\ket{1}+\alpha_{1}\beta_{0}\ket{1}\ket{0}+\alpha_{1}\beta_{1}\ket{1}\ket{1}
\end{aligned}
\end{equation}
in the four-dimensional Hilbert space. For a $n$-qubit system, if
we describe the $j$-th qubit as the state~$\ket{\psi_{j}}$, the
state of the whole system is tensor product of all subsystems $\ket{\Psi}=\ket{\psi_{1}}\otimes\cdots\otimes\ket{\psi_{j}}\otimes\cdots\otimes\ket{\psi_{n}}$
in the Hilbert space of dimension $D=2^{n}$. However, not all quantum
system can be expressed as the tensor product of its subsystems. This
kind of system is called entangled, and will be described in the next
section.\todo{This paragraph need to be polished.}

@@

\subsection{Deutsch Algorithm}

@@

Specifically, consider a one-qubit system with basis vectors $\ket{0}$
and~$\ket{1}$. In conventional quantum theory, there exists an infinite
number of states for a qubit of the form $\alpha_{0}\ket{0}+\alpha_{1}\ket{1}$,
with $\alpha_{0}$ and $\alpha_{1}$ elements of the underlying field
of complex numbers subject to the normalization condition $|\alpha_{0}|^{2}+|\alpha_{1}|^{2}=1$.
Moving to a finite field immediately limits the set of possible states
as the coefficients~$\alpha_{0}$ and~$\alpha_{1}$ are now drawn
from a finite set. In particular, in the field $\ff{2}=\{0,1\}$ of
booleans, there are exactly four possible vectors: the zero vector,
the vector~$\ket{0}$, the vector $\ket{1}$, and the vector $\ket{0}+\ket{1}=\ket{+}$.
Since the zero vector is considered non-physical, a one-qubit system
can be in one of only three states. The dynamics of these one-qubit
states is realized by any invertible linear map, i.e., by any linear
map that is guaranteed never to produce the zero vector from a valid
state. There are exactly $6$ such maps: 
\begin{equation}
\begin{aligned}\sigma_{0} & =\begin{pmatrix}1 & 0\\
0 & 1
\end{pmatrix}\,, & S & =\begin{pmatrix}1 & 0\\
1 & 1
\end{pmatrix}\,, &  & \begin{pmatrix}0 & 1\\
1 & 1
\end{pmatrix}\,,\\
\sigma_{x} & =\begin{pmatrix}0 & 1\\
1 & 0
\end{pmatrix}\,, & S^{\dagger} & =\begin{pmatrix}1 & 1\\
0 & 1
\end{pmatrix}\,, &  & \begin{pmatrix}1 & 1\\
1 & 0
\end{pmatrix}\,.
\end{aligned}
\end{equation}
This set of maps is clearly quite impoverished compared to the full
set of one-qubit unitary maps in conventional quantum theory. In particular,
it does not include the Hadamard transformation. However, this set
also includes non-unitary maps such as $S$ and $S^{\dagger}$ that
are not allowed in conventional quantum computation.

@@

The state is initialized to $\ket{0}\ket{\overline{0}}$, with $\ket{\overline{0}}=\ket{0}\ket{0}\cdots\ket{0}$,
i.e., the tensor product of $n$ $\ket{0}$ states. Applying the map
$S$ (defined in the previous section) to each qubit in the second
component of the state produces $\ket{0}\ket{\overline{+}}$ where
$\ket{\overline{+}}$ denotes the sequence $\ket{+}\ldots\ket{+}$
of length $n$. Applying $\uf$ to the entire state has no effect
since $U_{f}$ is the identity when $f$ is unsatisfiable. Applying
$S$ to each qubit in the second component of the state produces $\ket{0}\ket{\overline{0}}$.
Applying $S^{\dagger}$ to the first component leaves the state unchanged.
As the first component of the state is 0, applying the map $X_{0}$
(which is the identity) leaves the state unchanged. Applying $S^{\dagger}$
to the first component leaves the state unchanged. Measuring the state
will deterministically produce $\ket{0}\ket{\overline{0}}$.

>>>>

\begin{comment}
An $n$-qubit pure state in the conventional quantum theory (CQT)
is represented as a vector in the $2^{n}$-dimensional Hilbert space.
If we eliminate all symmetries, an irreducible state is actually a
point in the projective Hilbert space \cite{MosseriDandoloff2001,Jaeger2007}
also known as the complex projective space~$\CP{2^{n}-1}$ \cite{Hatcher2001,Bengtsson2007}.
These irreducible quantum states can also be classified as product
states and entangled states, and the latter one plays the essential
role for pure-state algorithms \cite{Mermin2007,Jaeger2007}.
\end{comment}
It has been 36 years since Feynman~\cite{Feynman1982Simulating}
proposed the idea about quantum computation. Although there are many
models for quantum computing, the circuit model is still one of the
most widespread models~\cite{544199} no matter whether in theoretical
study or in building a real quantum computer. Information in a quantum
circuit is stored in qubits. Each qubit is a two-dimensional quantum
system. For a two-qubit system, if the first and second qubit are
described as the state~$\ket{\psi}=\alpha_{0}\ket{0}+\alpha_{1}\ket{1}$
and $\ket{\phi}=\beta_{0}\ket{0}+\beta_{1}\ket{1}$, the state of
the whole system is their tensor product 
\begin{equation}
\begin{aligned} & \ket{\psi}\ket{\phi}=\ket{\psi}\otimes\ket{\phi}=\left(\alpha_{0}\ket{0}+\alpha_{1}\ket{1}\right)\otimes\left(\beta_{0}\ket{0}+\beta_{1}\ket{1}\right)\\
={} & \alpha_{0}\beta_{0}\ket{0}\otimes\ket{0}+\alpha_{0}\beta_{1}\ket{0}\otimes\ket{1}+\alpha_{1}\beta_{0}\ket{1}\otimes\ket{0}+\alpha_{1}\beta_{1}\ket{1}\otimes\ket{1}\\
={} & \alpha_{0}\beta_{0}\ket{0}\ket{0}+\alpha_{0}\beta_{1}\ket{0}\ket{1}+\alpha_{1}\beta_{0}\ket{1}\ket{0}+\alpha_{1}\beta_{1}\ket{1}\ket{1}
\end{aligned}
\end{equation}
in the four-dimensional Hilbert space. For a $n$-qubit system, if
we describe the $j$-th qubit as the state~$\ket{\psi_{j}}$, the
state of the whole system is tensor product of all subsystems 
\begin{equation}
\ket{\Psi}=\ket{\psi_{1}}\otimes\cdots\otimes\ket{\psi_{j}}\otimes\cdots\otimes\ket{\psi_{n}}=\ket{\psi_{1}}\cdots\ket{\psi_{j}}\cdots\ket{\psi_{n}}
\end{equation}
in the Hilbert space of dimension $D=2^{n}$. However, not all quantum
system can be expressed as the tensor product of its subsystems. This
kind of system is called entangled, and will be described in the next
section.\todo{This paragraph need to be polished.}

@@

\subsection{Quantum Circuits\label{subsec:Quantum-Circuits}}

@@

Specifically, consider a one-qubit system with basis vectors $\ket{0}$
and~$\ket{1}$. In conventional quantum theory, there exists an infinite
number of states for a qubit of the form $\alpha_{0}\ket{0}+\alpha_{1}\ket{1}$,
with $\alpha_{0}$ and $\alpha_{1}$ elements of the underlying field
of complex numbers subject to the normalization condition $|\alpha_{0}|^{2}+|\alpha_{1}|^{2}=1$.
Moving to a finite field immediately limits the set of possible states
as the coefficients~$\alpha_{0}$ and~$\alpha_{1}$ are now drawn
from a finite set. In particular, in the field $\ff{2}=\{0,1\}$ of
booleans, there are exactly four possible vectors: the zero vector,
the vector~$\ket{0}$, the vector $\ket{1}$, and the vector $\ket{0}+\ket{1}=\ket{+}$.
Since the zero vector is considered non-physical, a one-qubit system
can be in one of only three states. The dynamics of these one-qubit
states is realized by any invertible linear map, i.e., by any linear
map that is guaranteed never to produce the zero vector from a valid
state. There are exactly $6$ such maps, and their matrix representation
with respect to the standard basis are: 
\begin{equation}
\begin{aligned}\sigma_{0} & =\begin{pmatrix}1 & 0\\
0 & 1
\end{pmatrix}\,, & S & =\begin{pmatrix}1 & 0\\
1 & 1
\end{pmatrix}\,, &  & \begin{pmatrix}0 & 1\\
1 & 1
\end{pmatrix}\,,\\
\sigma_{x} & =\begin{pmatrix}0 & 1\\
1 & 0
\end{pmatrix}\,, & S^{\dagger} & =\begin{pmatrix}1 & 1\\
0 & 1
\end{pmatrix}\,, &  & \begin{pmatrix}1 & 1\\
1 & 0
\end{pmatrix}\,.
\end{aligned}
\label{eq:MQT-dynamics}
\end{equation}
For example, 
\begin{equation}
\begin{aligned}S\ket{0} & =\begin{pmatrix}1 & 0\\
1 & 1
\end{pmatrix}\begin{pmatrix}1\\
0
\end{pmatrix}=\begin{pmatrix}1\\
1
\end{pmatrix}=\ket{+}\,, & S\ket{+} & =\begin{pmatrix}1 & 0\\
1 & 1
\end{pmatrix}\begin{pmatrix}1\\
1
\end{pmatrix}=\begin{pmatrix}1\\
0
\end{pmatrix}=\ket{0}\,.\end{aligned}
\label{eq:MQT-dynamics-apply}
\end{equation}
This set of maps is clearly quite impoverished compared to the full
set of one-qubit unitary maps in conventional quantum theory. In particular,
it does not include the Hadamard transformation. However, this set
also includes non-unitary maps such as $S$ and $S^{\dagger}$ that
are not allowed in conventional quantum computation.

@@

\todo{This part use $\ket{\overline{a}}=\ket{a_{1}}\ldots\ket{a_{n}}$
while previous parts use $\ket{\Psi}=\ket{\psi_{1}}\cdots\ket{\psi_{j}}\cdots\ket{\psi_{n}}$.
Moreover, bar is heavily used in QIVPM discussion later, so maybe
not using bar here????}The state is initialized to $\ket{0}\ket{\overline{0}}$,
with $\ket{\overline{0}}=\ket{0}\ket{0}\cdots\ket{0}$, i.e., the
tensor product of $n$ $\ket{0}$ states. As Eq.~(\ref{eq:MQT-dynamics-apply}),
applying the map $S$ to each qubit in the second component of the
state produces $\ket{0}\ket{\overline{+}}$ where $\ket{\overline{+}}$
denotes the sequence $\ket{+}\ldots\ket{+}$ of length $n$. Applying
$\uf$ to the entire state has no effect since $U_{f}$ is the identity
when $f$ is unsatisfiable. Applying $S$ to each qubit in the second
component of the state produces $\ket{0}\ket{\overline{0}}$. Applying
$S^{\dagger}$ to the first component leaves the state unchanged.
As the first component of the state is $0$, applying the map $\sigma_{0}$
(which is the identity) leaves the state unchanged.\todo{Control-not
need to be defined and explained in Sec.~\ref{subsec:Quantum-Circuits}}
Applying $S^{\dagger}$ to the first component leaves the state unchanged.
Measuring the state will deterministically produce $\ket{0}\ket{\overline{0}}$.
